---
title: "Applied Economic Forecasting"
subtitle: "2. Exploring & Visualizing Time series"
date: "Spring 2020"
autosize: true
fontsize: 11pt
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 3.5
    highlight: tango
    theme: madrid
    colortheme: "default"
    fonttheme: "serif"
    toc: true
keep_tex: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=50),tidy=TRUE)
library(fpp2)
#options(width=50)
```

# Time series in R

## `ts` objects and `ts` function

\fontsize{12}{13}

A time series is stored in a `ts` object in R:

 - a list of numbers
 - information about times those numbers were recorded.

### Example

```{r tstable, echo=FALSE}
x <- c(123,39,78,52,110)
yr <- 2012:2016
knitr::kable(data.frame(Year=yr,Obs.=x), booktabs=TRUE)
```

```r
y <- ts(c(123,39,78,52,110), start=2012)
```

## `ts` objects and `ts` function

For observations that are more frequent than once per year, we need to add a `frequency` argument.

E.g., monthly data stored as a numerical vector `z`:

```r
y <- ts(z, frequency=12, start=c(2003, 1))
```

## `ts` objects and `ts` function

### `ts(data, frequency, start)`

\begin{tabular}{lrl}
\bf Type of data & \hspace*{1.95cm}\bf frequency                 & \bf start example\hspace*{0.25cm} \\
\midrule
Annual           &  1                                  & 1995\\
Quarterly        & 4                                  & c(1995,2)\\
Monthly          & 12                                 & c(1995,9)\\
Daily            & 7 \emph{or} 365.25                 & \emph{or} c(1995,234) \\
Weekly           & 52.18                             & c(1995,23)\\
Hourly           & 24 \emph{or} 168 \emph{or} 8,766  & 1\\
Half-hourly      & 48 \emph{or} 336 \emph{or} 17,532 & 1
\end{tabular}

## Let's Practice!!!

1. Set a seed as 10^3
2. Generate a random normal variable (`x`) with 200 observations, mean=75, and sd= 5
3. Declare x as a *quarterly* ts object ending December 2018 (`x.ts`)\label{step3}
4. Now repeat step \ref{step3} with weekly, monthly and annual frequencies (*How about using a loop?*)


## Australian GDP

\fontsize{12}{14}

```
ausgdp <- ts(x, frequency=4, start=c(1971,3))
```

  * Class: "ts"
  * Print and plotting methods available.

```{r, echo=TRUE}
head(ausgdp,30)
```

## Australian GDP

```{r, echo=TRUE, fig.height=4}
autoplot(ausgdp)
```

## Residential electricity sales

\fontsize{12}{14}\sf

```{r}
elecsales
```

## Class package

```
> library(fpp2)
```

This loads:

  * some data for use in examples and exercises
  * **forecast** package (for forecasting functions)
  * **ggplot2** package (for graphics functions)
  * **fma** package (for lots of time series data)
  * **expsmooth** package (for more time series data)

# Time plots

## Time plots

\small

```{r, echo=TRUE, fig.height=4}
autoplot(melsyd[,"Economy.Class"])
```

## Time plots

\small

```{r a10, echo=TRUE}
autoplot(a10) + ylab("$ million") + xlab("Year") + ggtitle("Antidiabetic drug sales")
```

## Let's Practice \setbeamercolor{}{bg=red}

1. Using the data from our earlier exercise: Plot `x.ts` at the monthly, quarterly and yearly frequencies.

2. Create plots of the following time series: `dole`, `bricksq`, `lynx`, `goog`
- Use `help()` to find out about the data in each series.
- For each of the plots, be sure to modify the axis labels and title.

## Are time plots best?

\fontsize{12}{14}\sf

```{r maxtemp}
autoplot(elecdaily[,"Temperature"]) + labs(x = "Week", y = "Max temperature")
```

## Are time plots best?

\fontsize{12}{14}\sf

```{r maxtemp2, warning=FALSE, message=FALSE}
qplot(time(elecdaily), elecdaily[,"Temperature"]) + labs(x = "Week", y = "Max temperature")
```

# Seasonal plots

## Seasonal plots

\footnotesize

```{r, echo=TRUE}
ggseasonplot(a10, year.labels=TRUE, year.labels.left=TRUE) +  ylab("$ million") +
  ggtitle("Seasonal plot: antidiabetic drug sales")
```

## Seasonal plots

  * Data plotted against the individual "seasons" in which the data were observed.  (In this case a "season" is a month.)
  * Something like a time plot except that the data from each season are overlapped.
  * Enables the underlying seasonal pattern to be seen more clearly, and also allows any substantial departures from the seasonal pattern to be easily identified.
  * In R: `ggseasonplot()`

## Seasonal polar plots

\small

```{r, fig.height=6, out.width="7.5cm"}
ggseasonplot(a10, polar=TRUE) + ylab("$ million")
```

## Seasonal subseries plots

\small

```{r, echo=TRUE}
ggsubseriesplot(a10) + ylab("$ million") +
  ggtitle("Subseries plot: antidiabetic drug sales")
```

## Seasonal subseries plots

  * Data for each season collected together in time plot as separate time series.
  * Enables the underlying seasonal pattern to be seen clearly, and changes in seasonality over time to be visualized.
  * In R: `ggsubseriesplot()`

## Quarterly Australian Beer Production

```{r}
beer <- window(ausbeer,start=1992)
autoplot(beer)
```

## Quarterly Australian Beer Production

```{r}
ggseasonplot(beer,year.labels=TRUE)
```

## Quarterly Australian Beer Production

```{r}
ggsubseriesplot(beer)
```

## Let's Practice!!!

The `arrivals` data set comprises quarterly international arrivals (in thousands) to Australia from Japan, New Zealand, UK and the US.

  - Use `autoplot()` and `ggseasonplot()` to compare the differences between the arrivals from these four countries.
  - Can you identify any unusual observations?

# Seasonal or cyclic?

## Time series patterns

Trend
  : pattern exists when there is a long-term increase or decrease in the data.

Seasonal
  : pattern exists when a series is influenced by seasonal factors (e.g., the quarter of the year, the month, or day of the week).

Cyclic
  : pattern exists when data exhibit rises and falls that are \emph{not of fixed period} (duration usually of at least 2 years).

## Time series components

### Differences between seasonal and cyclic patterns:

* seasonal pattern constant length; cyclic pattern variable length
* average length of cycle longer than length of seasonal pattern
* magnitude of cycle more variable than magnitude of seasonal pattern

## Time series patterns

\small

```{r}
autoplot(window(elec, start=1980)) +
  ggtitle("Australian electricity production") + labs(x = "Year", y = "GWh")
```

## Time series patterns

\small

```{r}
autoplot(bricksq) +
  ggtitle("Australian clay brick production") +  xlab("Year") + ylab("million units")
```

## Time series patterns

\small

```{r}
autoplot(hsales) +  ggtitle("Sales of new one-family houses, USA") + labs(x = "Year", y = "Total sales")
```

## Time series patterns

```{r}
autoplot(ustreas) + ggtitle("US Treasury Bill Contracts") +  labs(x = "Day", y = "price")
```

## Time series patterns
\small

```{r}
autoplot(lynx) +  ggtitle("Annual Canadian Lynx Trappings") +  xlab("Year") + ylab("Number trapped")
```

## Seasonal or cyclic?

### Differences between seasonal and cyclic patterns:

  * seasonal pattern constant length; cyclic pattern variable length
  * average length of cycle longer than length of seasonal pattern
  * magnitude of cycle more variable than magnitude of seasonal pattern
     
\bigskip

**The timing of peaks and troughs is predictable with seasonal data, but unpredictable in the long term with cyclic data.**

# Scatterplots

## Scatterplots: Relationship *Between* Time Series
\scriptsize
Scatterplots are commonly used tools for visualizing the relationship between variables. Take the text example of Electricty Demand and Temperature in Victoria, Autralia:

```{r tidy=TRUE}
qplot(x=Temperature, y=Demand, data=as.data.frame(elecdemand)) + labs(y = "Demand", x = "Temperature (C)")
```

## Scatterplots: Relationship *Between* Time Series {.plain}

\small
We notice that higher temperatures are associated with higher demand for electricity. 

### What if we would like to get a measure of the strength of this relationship?
- For this, we will use the **correlation coefficient**:

$$\rho_{xy} = \frac{\sum\limits_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum\limits_{i=1}^{n}(x_i-\bar{x})^2}\sqrt{\sum\limits_{i=1}^{n}(y-\bar{y})^2}} = \frac{\sum\limits_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sigma_x^2}\sqrt{\sigma_y^2}}, \quad -1 \le \rho_{xy} \le 1 $$

- Negative values indicate a negative *linear* relationship between `x` and `y`.

- Positive values indicate a positive *linear* relationship between `x` and `y`. 

## Scatterplot Matrices: Relationship *Between* Time Series {.plain}
\scriptsize
We can visualize the correlation between multiple series using a scatterplot matrix. In `R`, we will use the `ggpairs()` command from the `GGally` 

```{r, tidy=TRUE, fig.height=3.5, message=FALSE, keepaspectratio=TRUE}
#install.packages("GGally") #unmute when running for the first time
GGally::ggpairs(as.data.frame(visnights[,6:8]),columnLabels = c("Metro", "Central", "North"))
```

# Lag plots and autocorrelation

## Example: Beer production

```{r, echo=TRUE, fig.height=4.5, fig.width=8}
beer <- window(ausbeer, start=1992)
gglagplot(beer)
```

## Lagged scatterplots

  * Each graph shows $y_t$ plotted against $y_{t-k}$ for different values of $k$.
  * The autocorrelations are the correlations associated with these scatterplots.

## Autocorrelation

**Covariance** and **correlation**: measure extent of **linear relationship** between two variables ($y$ and $X$).

**Autocovariance** and **autocorrelation**: measure linear relationship between **lagged values** of a time series $y$.

We measure the relationship between:

  * $y_{t}$ and $y_{t-1}$
  * $y_{t}$ and $y_{t-2}$
  * $y_{t}$ and $y_{t-3}$
  * etc.

## Autocorrelation {.plain}

We denote the sample autocovariance at lag $k$ by $\gamma_k$ and the sample autocorrelation at lag $k$ by $\rho_k$.  Then define
\begin{align*}
\gamma_k &= \frac{1}{T}\sum_{t=k+1}^T (y_t-\bar{y})(y_{t-k}-\bar{y}) \\
\text{and}\qquad \rho_{k} &= \frac{\gamma_k}{\gamma_0}
\end{align*}

  * It is easy to see that $\gamma_0$ is the variance of y. Let `k = 0` then:
  
  $$ \gamma_0 = \frac{1}{T}\sum_{t=1}^T (y_t-\bar{y})(y_{t}-\bar{y}) = \frac{1}{T}\sum_{t=1}^T (y_t-\bar{y})^2$$
  
  * $\rho_1$ indicates how successive values of  $y$  relate to each other
  * $\rho_2$ indicates how  $y$ values two periods apart relate to each other
  * $\rho_k$ is \textit{almost} the same as the sample correlation between $y_t$ and $y_{t-k}$.

## Autocorrelation

\small
Results for first 9 lags for beer data:

\footnotesize

```{r, echo=FALSE}
beeracf <- matrix(acf(c(beer), lag.max=9,
                      plot=FALSE)$acf[-1,,1], nrow=1)
colnames(beeracf) <- paste("$\\rho_",1:9,"$",sep="")
knitr::kable(beeracf, booktabs=TRUE,
             align="c", digits=3,
             format.args=list(nsmall=3))
```

```{r beeracf, fig.height=2.5}
ggAcf(beer)
```

## Autocorrelation

  * $\rho_{4}$  higher than for the other lags. This is due to **the seasonal pattern in the data**: the peaks tend to be **4 quarters** apart and the troughs tend to be **2 quarters** apart.
  * $\rho_2$ is more negative than for the other lags because troughs tend to be 2 quarters behind peaks.
  * Together, the autocorrelations at lags 1, 2, \dots, make up the \emph{autocorrelation function} or ACF.
  * The plot is known as a **correlogram**

## ACF

```{r, fig.height=4, echo=TRUE}
ggAcf(beer,col="red")
```

## Trend and seasonality in ACF plots

- When data have a trend, the autocorrelations for small lags tend to be large and positive.
- When data are seasonal, the autocorrelations will be larger at the seasonal lags (i.e., at multiples of the seasonal frequency)
- When data are trended and seasonal, you see a combination of these effects.

## Aus monthly electricity production

```{r}
elec2 <- window(elec, start=1980)
autoplot(elec2)
```

## Aus monthly electricity production

```{r}
ggAcf(elec2, lag.max=48)
```

## Aus monthly electricity production

Time plot shows clear trend and seasonality.

The same features are reflected in the ACF.

  * The slowly decaying ACF indicates trend.
  * The ACF peaks at lags 12, 24, 36, \dots, indicate seasonality of length 12.

## Google stock price

```{r}
autoplot(goog,col="blue")
```

## Google stock price

```{r}
ggAcf(goog, lag.max=100)
```

## Let's Practice!!! 

We have introduced the following graphics functions:

  - `gglagplot`
  - `ggAcf`
  
Explore the following time series using these functions. Can you spot any seasonality, cyclicity and trend? What do you learn about the series?

  - `hsales`
  - `usdeaths`
  - `bricksq`
  - `sunspotarea`
  - `gasoline`

## Which is which?

```{r, fig.height=6, fig.width=12, echo=FALSE, warning=FALSE, out.width="11.5cm"}
tp1 <- autoplot(cowtemp) + xlab("") + ylab("chirps per minute") +
  ggtitle("1. Daily temperature of cow")
tp2 <- autoplot(USAccDeaths/1e3) + xlab("") + ylab("thousands") +
  ggtitle("2. Monthly accidental deaths")
tp3 <- autoplot(AirPassengers) + xlab("") + ylab("thousands") +
  ggtitle("3. Monthly air passengers")
tp4 <- autoplot(mink/1e3) + xlab("") + ylab("thousands") +
  ggtitle("4. Annual mink trappings")
acfb <- ggAcf(cowtemp, ci=0) + xlab("") + ggtitle("B") + ylim(-0.4,1)
acfa <- ggAcf(USAccDeaths, ci=0) + xlab("") + ggtitle("A") + ylim(-0.4,1)
acfd <- ggAcf(AirPassengers, ci=0) + xlab("") + ggtitle("D") + ylim(-0.4,1)
acfc <- ggAcf(mink, ci=0) + xlab("") + ggtitle("C") + ylim(-0.4,1)
gridExtra::grid.arrange(tp1,tp2,tp3,tp4,
                        acfa,acfb,acfc,acfd,nrow=2)
```

# White noise

## Example: White noise

```{r}
set.seed(14567)
wn <- ts(rnorm(36))
autoplot(wn) + ggtitle("White Noise")
```

## Example: White noise

```{r, results='asis', echo=FALSE,fig.height=2}
ggAcf(wn)
wnacf <- matrix(acf(c(wn), lag.max=7,
                      plot=FALSE)$acf[-1,,1], nrow=1)
colnames(wnacf) <- paste("$\\rho_",1:7,"$",sep="")
knitr::kable(wnacf, booktabs=TRUE,
             align="c", digits=3,
             format.args=list(nsmall=3))

```

We expect each autocorrelation to be close to zero.

## \large Sampling distribution of autocorrelations

Sampling distribution of $\rho_k$ for white noise data is asymptotically N(0,$\frac{1}{T}$).

  *  95% of all $\rho_k$ for white noise must lie within $\pm \frac{1.96}{\sqrt{T}}$.
  * If this is not the case, the series is probably not WN.
  * Common to plot lines at $\pm 1.96/\sqrt{T}$ when plotting ACF.
These are the \textcolor{orange}{\textbf{\emph{critical values}}}.

## Autocorrelation

\structure{Example:}

```{r results='asis',fig.height=2}
ggAcf(wn)
```

$T=36$ and so critical values at $\pm \frac{1.96}{\sqrt{36}} = \pm 0.327$.

All autocorrelation coefficients lie within these limits. Data cannot be distinguished from white noise.

## Example: Pigs slaughtered

\scriptsize
```{r, fig.height=3,results='asis'}
pigs2 <- window(pigs, start=1990)
autoplot(pigs2) +
  xlab("Year") + ylab("thousands") + 
  ggtitle("Number of pigs slaughtered in Victoria")
```

## Example: Pigs slaughtered

```{r}
ggAcf(pigs2)
```

## Example: Pigs slaughtered

Monthly total number of pigs slaughtered
in the state of Victoria, Australia, from January 1990 through August 1995.
(Source: Australian Bureau of Statistics.)

  * Difficult to detect pattern in time plot.
  * ACF shows some significant autocorrelation at lags 1, 2, and 3.
  * $\rho_{12}$ relatively large although not significant.  This may indicate
some slight seasonality.


These show the series is **not a white noise series**.

## Let's Practice!!!

1. Compute the daily changes in the Google stock price (`ddgoog`).
2. Does `ddgoog` look like white noise?
    - Present a time plot of `ddgoog`
    - Present the ACF plot

## Let's Practice (Solution)
\structure{Manually:}
\scriptsize
``` {r}
#initialize the ddgoog variable to be a object with NAs
ddgoog <- NA

# We can use the loop to do the changes calculations
# Remember with differencing we lose the first obs so we start the counter at 2
for (i in 2:length(goog)){ 
ddgoog[i] <- goog[i] - goog[i-1] # Store results in ddgoog at position [i]
}
#Declare as a ts and drop first obs since it's NA
ddgoog <- ts(ddgoog[-1],start=c(2))
round(head(ddgoog),3)

```
---
\scriptsize

```{r, tidy = TRUE, fig.height=2.15, out.width="11.5cm"}
autoplot(ddgoog) + ggtitle("Daily Chances in Google Stock Prices") + labs(y = "Daily Changes", x = "")

ggAcf(ddgoog)
```

## Let's Practice (Solution) {.plain}
\structure{Using the fpp2 package:}
\scriptsize

```{r tidy= TRUE, fig.height=2.5,out.width="11.5cm"}
dgoog <- diff(goog)
round(head(dgoog),3)
autoplot(dgoog)+ggtitle("Daily Chances in Google Stock Prices") + labs(y= "Daily Changes", x= "")
```

## Let's Practice (Solution) {.plain}
\Large \center
__So, what is your conclusion?__
