---
title: "Applied Economic Forecasting"
subtitle: 'Tutorial on R Studio III'
#author: Shamar L. Stewart
#institute: |
#  | Department of Agricultural & Applied Economics
#  | Virginia Tech
date: 'Spring 2020'
autosize: true
fontsize: 11pt
header-includes:
   - \usepackage{xcolor}

output:
  beamer_presentation:
    keep_tex: FALSE
    theme: "madrid"
    colortheme: "default"
    fonttheme: "serif"
    slide_level: 2
    # toc: TRUE
    includes:
      in_header: 
        - ../mystyle.tex

---

<!-- In class, you can download this set of code [here](R_intro_III_Codes.R) -->
  
## Data types in `R`

Here we discuss various types of data in `R`. `R` works with numerous data types. Some of the most basic types to get started are:

  1. **Decimal values** like 4.5 are called numerics.
  2. **Natural numbers**  like 4 are called integers. Integers are also numerics.
  3. **Boolean values**  (TRUE or FALSE) are also called logical.
  4. **Text**  (or string) values are called characters.



You can check the type of data by using `class()`. 
```{r}
  x <- "Melody to Old Town Road!"
  class(x)
```

---

## {.plain}

```{r}
    x <- c("TRUE", "FALSE")
    x <- as.logical(x) #Declare the data type
    class(x)
    x <- 1:20
    x %% 4 #x mod 4
    x %% 4 == 0
    class(x %% 4 == 0)
```


## A collection of data points

__Vectors	__

A vector is the most common and basic data type in R, and is pretty much the workhorse of R. A vector is composed by a series of values, which can be either numbers or characters. We can assign a series of values to a vector using the c() function. 

Here `c()` stands for **concatenate** or **combine**. 

---

### Some basic examples

```{r}
(v <- c(1, 2, 3, 4))
(v <- 1:4)
(v <- seq(from = 0, to = 0.5, by = 0.1))

#A vector can also contain characters:	
(v_colors <- c("blue", "yellow", "light green")	)

```


## Subsetting vectors (Indexing/reassigning elements)

```{r}
v_colors[2]         
v_colors[c(1, 3)]   
(v_colors[2:3]  <- c("red", "purple")	)
j <- c(-1,-2,-3)
x[j]
```


## Conditional subsetting

Another common way of subsetting is by using a logical vector. `TRUE` will select the element with the same index, while `FALSE` will not. Typically, these logical vectors are not typed by hand, but are the output of other functions or logical tests such as:

```{r , tidy=TRUE, tidy.opts=list(width.cutoff=50)}
x <- 100:110
# returns TRUE or FALSE depending on which elements that meet the condition
x >105 
select <- x > 105
x[x > 105]
```


---

You can combine multiple tests using `&` (both conditions are true, `AND`) or `|` (at __least one__ of the conditions is true, `OR`):

x is between 103 and 106

```{r}
x[x >= 103 & x <= 106]
```

x is greater than 103 but less than or equal to 106

```{r}
x[x <= 106 & x > 103] # order of subsetting does not matter here!

```

x is less than 103 `or` greater than 106
```{r}
x[x >= 106 | x < 103] 

```

---

Sometimes we will need to search for certain strings in a vector. With mutlpile conditions, it becomes difficult to use the "OR" operator `|`. The function `%in%` allows you to test if any of the elements of a search vector are found:

```{r}
animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat

animals %in% c("rat", "cat", "dog", "duck", "goat")
animals[animals %in% c("rat", "cat", "dog", "duck", "goat")]
```

### Additional Resources

If you would like to get a better grasp on subsetting vectors, please feel free to try out [R Intro on Datacamp](https://campus.datacamp.com/courses/free-introduction-to-r). I would highly reccommend this!


## Names of a vector

Let's say that we want to know which color robe each of 3 patients is wearing, we can assign names to the vector of colors.
```{r}
v_colors
names(v_colors) <- c("Thomas", "Liz", "Tucker")
v_colors
```

## Algebraic Operations of Vectors

```{r}
x <- c(1,2,3)
y <- c(4,5,6)
# component-wise addition
x+y
# component-wise multiplication
x*y
# What happens to the following
y^x # or y**x
```


## Repeating Vector in `R`

```{r, warning=FALSE}
# Would this work?
c(1,2,3,4) + c(1,2)
# Would this work?
c(1,2,3) + c(1,2)
```

*Why the weird results?*

- When you are adding vectors of unequal size, if the long one is a multiple of the short one, `R` automatically repeats the short one to fill in the operation. 

```{r}
2*c(1,2,3)
```

## Matrix {.plain}
**Create a new matrix**
```{r}
(matrix<-matrix(1:16, nrow = 4, byrow = TRUE))
```

Note that `:` means every number from 1 to 4. In the `matrix()` function:

  1. The first argument is the collection of elements that `R` will arrange into the rows and columns of the matrix. Here, we use 1:16 which is a shortcut for c(1, 2, 3, 4, ... 16).
  2. The argument `byrow` indicates that the matrix is filled by the rows. If we want the matrix to be filled by the columns, we use `byrow = FALSE`.
  3. The argument `nrow` indicates that the matrix should have 4 rows.	

## Selection of Matrix Elements

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
#Extract element in the first row and second column
matrix[1,2] 
#Extract the entire first row and second columns 
matrix[,1:2] 
```


## Assign dimension names to Matrix

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
	rownames(matrix) <- c("Yes", "No", "Perhaps", "Maybe")
	colnames(matrix) <- c("Apple", "Pear", "Banana", "Grapes")
	matrix
```

## Dimension of a matrix vs vector

```{r}
x <- c(1,2,3)
matrix<-matrix(1:4, byrow = TRUE, nrow = 2)
length(x)
length(matrix)
dim(matrix)
dim(x)
```


## Lists

`R` doesn’t like vectors to have different types: `c(TRUE, 1, "Frank")` becomes `c("TRUE", "1", "Frank")`. But storing objects with different types is absolutely fundamental to data analysis. `R` has a different type of object besides a vector used to store data of different types side-by-side: a list:

```{r}
c(TRUE, 1, "Frank")
x <- list(TRUE, 1, "Frank")

```

Many different things not necessarily of same length can be put together. 

```{r}
x <- list(c(1:5), c("a", "b","c"), c(TRUE, FALSE), c(5L, 6L))
```

	
## Dataframes

- Data frames are like spreadsheet data, rectangular with rows and columns. 
- Ideally each row represents data on a single observation and each column contains data on a single variable, or characteristic, of the observation. 
- It represents the data in  a tabular format where the columns are vectors that all have the same length. Because columns are vectors, each column must contain a single type of data (e.g., characters, integers, factors). 
- We can open a data viewer window to see the contents of `R`'s `iris` data frame by typing

```{r, eval=FALSE}
View(iris)
```


## Create a data frame

Data frame with Harry Potter characters

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
	name <- c("Harry", "Ron", "Hermione", "Hagrid", "Voldemort")	
	height <- c(176, 175, 167, 230, 180)	
	gpa <- c(3.4, 2.8, 4.0, 2.2, 3.4)	
	df_students <- data.frame(name, height, gpa)  		
	df_students	
```

--- 

Alternative way of creating DF

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=50)}
	df_students <- data.frame(name = c("Harry", "Ron", "Hermione", "Hagrid",
					   "Voldemort"), 	
				  height = c(176, 175, 167, 230, 180), 	
				  gpa = c(3.4, 2.8, 4.0, 2.2, 3.4))	
	df_students
```


## Adding variable

```{r}
	df_students$good <- c(1, 1, 1, 1, 0)   	
	df_students	
```


## Features of the DF

```{r, eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=50)}
	dim(df_students)  		
	df_students[2, 3]               #Ron's GPA
	df_students$gpa[2]              #Ron's GPA
	df_students[5, ]                #get row 5
	df_students[3:5, ]              #get rows 3-5
	df_students[, 2]                #get column 2 (height)
	df_students$height              #get column 2 (height)
	df_students[, 1:3]              #get columns 1-3
	df_students[4, 2] <- 255        #reassign Hagrid's height
	df_students$height[4] <- 255    #same thing as above
	df_students		
```

