---
title: "Applied Economic Forecasting"
subtitle: 'Tutorial on R Studio IV'
#author: Shamar L. Stewart
#institute: |
#  | Department of Agricultural & Applied Economics
#  | Virginia Tech
date: 'Spring 2020'
autosize: true
fontsize: 11pt
header-includes:
   - \usepackage{xcolor}

output:
  beamer_presentation:
    keep_tex: FALSE
    theme: "madrid"
    colortheme: "default"
    fonttheme: "serif"
    slide_level: 2
    # toc: TRUE
    includes:
      in_header: 
        - ../mystyle.tex
---

## Creating Directory and Set working directory

You can use the 
```{r, eval = FALSE}
getwd()
``` 
command to obtain the current directory `R` is using.

It is good practice to set the working directory location to where the files and data are stored. 

- Consider setting your working directory to a folder called AAEC4984 or AAEC5984 on your desktop (for example).

## Creating Directory and Set working directory

### Windows
```{r, tidy=TRUE, tidy.opts=list(width.cutoff=50),eval = FALSE}
  setwd("C:/users/[your user name]/Desktop/AAEC4984/")
  # OR
  setwd("C:\\users\\[your user name]\\Desktop\\AAEC4984\\") 
  # notice the double backslashes
```

### Mac
```{r, eval = FALSE}
  setwd("~/Desktop/AAEC4984")
```

---

- To check whether the wd is correct, we again use

```{r, eval = FALSE}
getwd()
```

- To obtain a list of the names of files or folders in the working directory, we can use

```{r, eval = FALSE}
  dir()
```

- To create a new folder in your directory we can use 
```{r, eval = FALSE}
  dir.create("[Folder name]")
```  

## Importing data

`R` allows us to import several file types. I will discuss 3 that we are most likely to use in this course.

1. Text files: Data sometimes come with headers (the first row is variable names, not actual data!) You need to tell R that!
 
```{r, eval= FALSE}
  textdata<-read.table("examples/hogsdata.txt",header=T)
```
  
2. CSV files
```{r,eval=FALSE}
  csvdata<-read.csv("examples/hogsdata.csv",header=T)
```

3. xlsx files (requires openxlsx package)
```{r, eval= FALSE}
  xlsxdata<-read.csv("examples/hogsdata.xlsx", ... )
```


## Using Functions & Packages

Functions are “canned scripts” that automate more complicated sets of commands including operations assignments, etc. For the purpose of this course, we will use a lot of functions that are built both in base R (that is, they are predfined) or available through R packages (discuss below).

A function usually takes one or more inputs called *arguments*, and often (but not always) return a *value*.

---

Consider for example, taking the average of a set of random numbers (x). 

```{r, results='markup'}
set.seed(124) 
x <- rnorm(6) * 100
(round(x, digits=2)) # round function => 2dp
```

If we were to do this manually, we would:

1. Sum up the values 
```{r}
sumx <- sum(x)
```  

2. Get the number of observations
```{r}
nx <- length(x)
```  

3. Divide sum by total number of observations

```{r}
meanx <- sumx/nx
```
 
 ---
 
 Using `R`'s built in `mean` functions we can do all three steps internally and cross check against our manual calculations.

```{r}
mean(x)
meanx == mean(x) # cross validation
```

## Installing Packages

Since R is an Open Source software program, thousands of people contribute to the software. They do this by writing commands (called functions) to make a particular analysis easier, or to make a graphic prettier.

When you download R, you get access to a lot of functions that we will use. However the other user-written packages we use for our analyses will make our lives much easier.

For example, though we can use the `plot` command for standard graphics, you will quickly see that we can get much better looking time graphs using the `fpp2` package (which also uses another package called ggplot2).


## Installing Packages

To install the `fpp2` package, we can use the command
```{r, eval=FALSE}
install.packages("fpp2")
```
We will need to install a package only once in `R`.

Now that you have the `fpp2` package installed, we can check to see if it is in use
```{r eval=FALSE}
search()
```

Lastly, in order to use the package, we will need to load the library 
```{r eval=TRUE, message= FALSE}
library(fpp2)
```

## Using libraries {.plain}

The `fpp2` package contains a number of useful datasets. One such data set is `h02`. 

Use the `help()` function to get a decription of this data. Try
```{r,eval= FALSE}
help(h02)
```

Now let us create a nice plot of the `h02` data

```{r , fig.height= 2.5}
autoplot(h02)
```

Let us leave it there for now!
